Introduction

### 2024-04-11 Sections/Introduction.qmd
finish intro

---
#title: Introduction
format:
  html: default
  hugo-md:
    echo: false
    output: false
    html-math-method: mathjax
    output-file: d_intro_hugo.md
  gfm:
    echo: true
    output-file: d_into_gfm.md
---

@@

In the domain of category learning, the constant vs. varied comparison is much less suitable. Instead, researchers will 
typically employ designs where all training groups encounter numerous stimuli, but one group experiences a greater number of unique exemplars [@wahlheimMetacognitiveJudgmentsRepetition2012; @nosofskyModelguidedSearchOptimal2019; @doyleMetacognitiveMonitoringCategory2016; @hoschPriorExperienceVariability2023; @brunsteinPreparingNoveltyDiverse2011], or designs where the number of unique exemplars is held constant, but one group trains with items that are more dispersed, or spread out across the category space [@posnerGenesisAbstractIdeas1968;  @homaCategoryBreadthAbstraction1976; @huHighvariabilityTrainingDoes2024; @bowmanTrainingSetCoherence2020; @maddoxStimulusRangeDiscontinuity2011].

Much of the earlier work in this sub-area trained subjects on artificial categories, such as dot patterns [@homaCategoryBreadthAbstraction1976; @posnerGenesisAbstractIdeas1968], where more varied or distorted training examples were often shown to produce superior generalization when categorizing novel exemplars. More recently, researchers have also begun to utilize more realistic stimuli in their experiments. @wahlheimMetacognitiveJudgmentsRepetition2012 conducted one such study. In a within-participants design, participants were trained on bird categories with either high repetitions of a few exemplars, or few repetitions of many exemplars. Across four different experiments, which were conducted to address an unrelated question on metacognitive judgements, the researchers consistently found that participants generalized better to novel species following training with more unique exemplars (i.e. higher variability), while high repetition training produced significantly better performance categorizing the specific species they had trained on. A variability advantage was also found in the relatively complex domain of rock categorization [@nosofskyModelguidedSearchOptimal2019]. For 10 different rock categories, participants were trained with either many repetitions of 3 unique examples of each category, or few repetitions of 9 unique examples, with an equal number of total training trials in each group (the design also included 2 other conditions less amenable to considering the impact of variation). The high-variability group, trained with 9 unique examples, showed significantly better generalization performance than the other conditions. Moreover, the pattern of results in this study could be nicely accounted for by an extended version of the Generalized Context Model.

The studies described thus far have studied the benefits of variability by exposing participants to a greater or lesser number of distinct examples during training. A distinct sub-literature within the category learning domain has focused much less on benefits derived from varied training, instead emphasizing how increased variability during the learning of a novel category influences how far the category boundary will then be generalized. The general approach is to train participants on examples from two categories, with the examples from one of the categories being more dispersed than the other. Participants are then tested with novel items located within ambiguous regions of the task space which allow the experimenters to assess whether the difference in variability influences how far participants generalize the category boundaries.

@cohenCategoryVariabilityExemplar2001 trained subjects on two categories, one with much more variability than the other. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into a category with greater variation. Moreover, this effect was at odds with the predications of the baseline version of the GCM, thus providing some evidence that training variation may at least sometimes induce effects that cannot be entirely accounted for by exemplar-similarity accounts.

Further observations consonant with the results of @cohenCategoryVariabilityExemplar2001  have since been observed in numerous investigations [@hahnEffectsCategoryDiversity2005; @perlmanFurtherAttemptsClarify2012;
@sakamotoPuttingPsychologyBack2008; @hsuEffectsGenerativeDiscriminative2010]. The results of @sakamotoPuttingPsychologyBack2008 are noteworthy. They first reproduced the basic finding of participants being more likely to categorize an unknown middle stimulus into a training category with higher variability. In a second experiment, they held the variability between the two training categories constant and instead manipulated the training sequence, such that the examples of one category appeared in an ordered fashion, with very small changes from one example to the other (the stimuli were lines that varied only in length), whereas examples in the alternate category were shown in a random order and thus included larger jumps in the stimulus space from trial to trial. They found that the middle stimulus was more likely to be categorized into the category that had been learned with a random sequence, which was attributed to an increased perception of variability which resulted from the larger trial to trial discrepancies.

The work of @hahnEffectsCategoryDiversity2005, is also of particular interest to the present discussion. Their experimental design was similar to previous studies, but they included a larger set of testing items which were used to assess generalization both between the two training categories as well as novel items located in the outer edges of the training categories. During generalization testing, participants were given the option to respond with "neither", in addition to responses to the two training categories. The "neither" response was included to test how far away in the stimulus space participants would continue to categorize novel items as belonging to a trained category. Consistent with prior findings, high-variability training resulted in an increased probability of categorizing items in between the training categories as belong to the high variability category. Additionally, participants trained with higher variability also extended the category boundary further out into the periphery than participants trained with a lower variability category were willing to do. The authors then used the standard GCM framework to compare a variety of similarity-based models to account for their results. Of particular interest are their evaluations of a category response bias parameter, and a similarity scaling parameter. A model fit improvement when the response bias parameter is allowed to vary between the high-variability and low-variability trained groups is taken to suggest a simple bias for responding with one of the trained categories over the other. Alternatively, an improvement in fit due to a separate similarity scaling parameter may reflect the groups being differentially sensitive to the distances between stimuli. No improvement in model fit was found by allowing the response-bias parameter to differ between groups, however the model performance did improvement significantly when the similarity scaling parameter was fit separately. The best fitting similarity-scaling parameters were such that the high-variability group was less sensitive to the distances between stimuli, resulting in greater similarity values between their training items and testing items. This model accounted for both the extended generalization gradients of the varied particpants, and also for their poor performance in a recognition condition. Additional model comparisons suggested that this similarity rescaling applied across the entire stimulus space, rather than to the high variability category in particular.

Variability effects have also been examined in the higher-level domain of how learners acquire novel concepts, and then instantiate (rather than merely recognize) that concept in untrained contexts [@braithwaiteEffectsVariationPrior2015]. This study trained participants on problems involving the concept of sampling with replacement (SWR). Training consisted of examples that were either highly similar in their semantic context (e.g. all involving people selecting objects) or in which the surface features were varied between examples (e.g. people choosing objects AND objects selected in a sequence). The experimenters also surveyed how much prior knowledge each participant had with SWR. They found that whether variation was beneficial depended on the prior knowledge of the participants -- such that participants with some prior knowledge benefited from varied training, whereas participants with minimal prior knowledge performed better after training with similar examples. The authors hypothesized that in order to benefit from varied examples, participants must be able to detect the structure common to the diverse examples, and that participants with prior knowledge are more likely to be sensitive to such structure, and thus to benefit from varied training. To test this hypothesis more directly, the authors conducted a 2nd experiment, wherein they controlled prior knowledge by exposing some subjects to a short graphical or verbal pre-training lesson, designed to increase sensitivity to the training examples. Consistent with their hypothesis, participants exposed to the structural sensitivity pre-training benefited more from varied training than the controls participants who benefited more from training with similar examples.

Variability has also been examined within the realm of language learning. A particularly impressive study is that of [@perryLearnLocallyThink2010]. In nine training sessions spread out over nine weeks infants were trained on object labels in a naturalistic play setting. All infants were introduced to three novel objects of the same category, with participants in the tight condition being exposed to three similar exemplars of the category, and participants in the varied condition being exposed to three dissimilar objects of the same category. Importantly, the similarity of the objects was carefully controlled for by having a separate group of adult subjects provide pairwise similarity judgements of the category objects prior to the study onset. Multidimensional scaling was then performed to obtain the coordinates of the objects psychological space, and out of the 10 objects for each category, the 3 most similar objects were selected for the tight group and the three least similar objects for the varied group, with the leftover four objects being retained for testing. By the end of the nine weeks, all of the infants had learned the labels of the training objects. The varied group demonstrated superior ability to correctly generalize the object labels to untrained exemplars of the same category, a pattern consistent with much of the existing literature. More interesting was the superior performance of the varied group on a higher order generalization task -- such that they were able to appropriately generalize the bias they had learned during training for attending to the shape of objects to novel solid objects, but not to non-solids. The tight training group, on the other hand, tended to overgeneralize the shape bias, leading the researchers to suggest that the varied training induced a more context-sensitive understanding of when to apply their knowledge.

Of course, the relationship between training variability and transfer is unlikely to be a simple function wherein increased variation is always beneficial. Numerous studies have found null, or in some cases negative effects of training variation [@deloshExtrapolationSineQua1997; @sinkeviciuteRoleInputVariability2019; @wrisbergVariabilityPracticeHypothesis1987], and many more have suggested that the benefits of variability may depend on additional factors such as prior task experience, the order of training trials, or the type of transfer being measured [@bernikerEffectsTrainingBreadth2014; @braithwaiteEffectsVariationPrior2015; @hahnEffectsCategoryDiversity2005; @lavanEffectsHighVariability2019; @northEffectConsistentVaried2019; @sadakataIndividualAptitudeMandarin2014; @zamanPerceptualVariabilityImplications2021].

## Variability Literature Review

An early and influential work on the influence of variability on category learning is that of @posnerGenesisAbstractIdeas1968. In an ambitious attempt to address the question of how category information is represented, the authors trained participants to categorize artificial dot patterns, manipulating whether learners were exposed to examples clustered close to the category prototypes (e.g. a low variability condition), or spread further away from the prototype (the varied-training group). It should be noted that both groups in this study were trained with the same number of unique instances and the manipulated difference was how spread out the instances were. The authors claim based on prior experiments using the same stimuli, that the training stimuli for the varied group were at least as far away from the testing stimuli as the training stimuli of the less-varied group. The authors interpreted their findings as evidence for the extraction of an abstraction or schema that is extracted and stored, and then over time becomes more likely to be the reference point from which generalization occurs, given that specific instances are thought to decay at a faster rate than prototypes or schema. The Posner and Keele study has been extremely influential and continues to be cited in contemporary research as clear evidence that schema abstraction underlies the benefits of varied training. It's also referenced as a key influence in the development of "Schema Theory of Motor Learning" [@schmidtSchemaTheoryDiscrete1975], which in turn influenced decades of investigations on the potential benefits of varied training in motor skill learning. However, the classic Posner & Keele study despite being far more carefully designed than many subsequent studies, and despite being a relative rarity in explicitly discussing and attempting to control for potential confounds of similarity between groups, may nevertheless be emblematic of a common issue in many investigations of the effects of varied training on learning. The problem with Posner & Keele's conclusion was demonstrated clearly almost 3 decades later [@palmeriCentralTendenciesExtreme2001], when researchers conducting a near replication of the original study also collected similarity judgements following training and performed multidimensional scaling analysis. Rather than being in the middle of the training stimuli as was the case in the physical stimuli space, the psychological representation of the prototype was shown reside at an extreme point, and generalization patterns by participants that would have seemed to warrant the learning of a prototype were then easily accounted for with only the assumption that the participants encoded instances. One of the primary concerns of the present paper is that many of the studies which purport to explain the benefits of variation via prototypes, schemas, or other abstractions, are often overlooking the potential of instance based similarity accounts.

## Category learning

## Complications to the influence of variability

### Sequence Effects

A necessary consequence of varied training is that participants will have the experience of switching from one task condition to another. The number of switches can vary greatly, with the two extremes being varied participants completing all of their training trials in one before switching to the next condition (blocked sequencing), or if they alternate between conditions on a trial by trial basis (random/intermixed/interleaved sequencing). Not long after the initial influx of schema-theory inspired studies testing the benefits of variability hypothesis was shown that the influence of varied training might interact with the type of training sequence chosen by the experimenter  [@sheaContextualInterferenceEffects1979]. In this seminal study, both groups of training subjects trained with the same number of trials of three separate movement patterns. A blocked group that completed all of their trials with one sequence before beginning the next sequence, and a random group that trained with all three movement patterns interspersed throughout the course of training. Participants were also randomly assigned to retention testing under either blocked or random sequence conditions, thus resulting in all four training-testing combinations of blocked-blocked; blocked-random; random-blocked; random-random. There was some effect of sequence context, such that both groups performed better when the testing sequence matched their training sequence. However, the main finding of interest was the advantage of random-training, which resulted in superior testing performance than blocked training regardless of whether the testing stage had a blocked or random sequence, an effect observed both immediately after training, and in a follow up test ten days after the end of training.

Prior to the influential @sheaContextualInterferenceEffects1979 study, studies investigating the benefits of variability hypothesis had utilized both blocked and random training schedules, often without comment or justification. It was later observed [@leeInfluencePracticeSchedule1985] that positive evidence for benefits of varied training seemed more likely to occur for studies that utilized random schedules. The theoretical basis of such studies was invariably an appeal to Schmidt's schema theory; however schema theory made no clear predictions of an effect of study sequence on retention or generalization, thus prompting the need for alternate accounts. One such account, the elaborative processing account [@sheaContextEffectsMemory1983], draws on the earlier work [@battigFacilitationInterference1966] and argues that randomly sequencing conditions during training promotes comparison and contrastive processes between those conditions, which result in a deeper understanding of the training task than could arise via blocked sequencing. Supporting evidence for elaborative processing comes in the form of random-sequence trained subjects self-reporting more nuanced mental representations of movement patterns following training [@sheaContextEffectsMemory1983], and by manipulating whether subjects are able to perform comparisons during training [@wrightContributionElaborativeProcessing1992]. An alternative, though not incompatible account suggests that the benefits of random-sequencing are a result of such sequences forcing the learner to continually reconstruct the relevant motor task in working memory [@leeLocusContextualInterference]. Blocked training, on the other hand, allows the learner to maintain the same motor task in short term memory without decay for much of the training which facilitates training performance, but hinders ability to retrieve the appropriate motor memory in a later testing context. A much more recent study [@chuaPracticeVariabilityPromotes2019], replicates the standard findings of an advantage of varied training over constant training (expt 1, bean-bag throwing task), and of random training over blocked training (expt 2 & 3, bean-bag throwing & golf putting). The novelty of this study is that the experimenters queried subjects about their attentional focus throughout the training stage. In all three experiments varied or random trained-subjects reported significantly greater external attention (e.g. attending to the target distance), and constant or blocked subjects reported more internal attention (e.g. posture or hand position). The authors argue that the benefits of varied/random training may be mediated by changes in attentional focus, however the claims made in the paper seem to go far beyond what can be justified by the analyses reported -- e.g. the increased external focus could be a simple byproduct of varied training. A stronger form of evidence that was not provided may have been to use multiple regression analyses to show that the testing advantage of the varied/random groups over the constant/blocked groups could be accounted for by the differences in self-reported attentional focus.

## Other task and participant effects

Of course, the effects of varied training, and different training sequences, are likely to be far more complex than simply more varied training being better than less, or random training being better than blocked. Null effects of both manipulations have been reported [see @magillReviewContextualInterference1990; @vanrossumSchmidtSchemaTheory1990 for reviews], and a variety of moderators have emerged. In one of the earlier examples of the complex relationship between study sequence and learning [@delreyEffectsContextualInterference1982], experimenters recruited participants who self-reported either large amounts, or very little experience with athletic activities, and then trained participants on a coincident timing task under with either a single constant training velocity, or with four training velocities under either blocked, or random training sequence conditions - resulting in six experimental conditions: (athlete vs. non-athlete) x (constant vs. blocked vs. random training). Athlete participants had superior performance during training, regardless of sequence condition, and training performance was superior for all subjects in the constant group, followed by blocked training in the middle, and then random training resulting in the worst training performance. Of greater interest is the pattern of testing results for novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received random training, followed by blocked, and then constant training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training. A similar pattern was later observed in a golf-putting training study, wherein participants who had some prior golf experience benefited most from random-sequenced training, and participants with no golf experience benefited most from blocked training  [@guadagnoliRelationshipContextualInterference1999]. More recently, the same pattern was observed in the concept learning literature [@braithwaiteEffectsVariationPrior2015 expt 1.]. This study trained participants on a mathematical concept and found that participants who self-reported some prior experience with the concept improved more from pre-test to post-test after training with varied examples, while participants who reported no prior experience showed greater gains following training with highly similar examples.

In addition to the influence of prior experiences described above, ample evidence also suggests that numerous aspects of the experiment may also interact with the influence of variation. One important study examined the impact of the amount of training completed in a force production task [@sheaContextualInterferenceContributions1990]. This study employed a typical blocked vs. random training procedure, but with the additional manipulation of separate groups receiving 50, 200, or 400 total training trials. For the group that received only 50 training trials retention was best when training had been blocked. However, for the conditions that received 200 or 400 training trials the pattern was reversed, with random training resulting in superior retention than blocked training. These results were taken to suggest that the benefits of randomization may take time to emerge. Another experimental factor shown to interact with training sequence is the complexity of the training task [@albaretDifferentialEffectsTask1998]. In addition to random or blocked training, participants in this study were assigned to train on a drawing task at one of three different levels of complexity (reproducing from memory shapes composed of two, three or four components). On a transfer task 48 hours after the completion of training, only participants trained at the lower levels of task complexity (2 or 3 components) showed superior performance to the blocked condition. The authors suggest that the benefits of random sequencing, thought to arise from more elaborate cognitive processing, or the necessity of continually recalling task information from long term into short term memory, are more likely to be obscured as the complexity of the task forces the blocked participants to also engage in such processes.

A final important influence of particular relevance to the practice-sequence literature concerns the exact structure of "random" sequencing. Although the term random is commonly used for convenience, experimenters do not typically leave the order of training entirely up to chance. Rather, the training sequence is often constrained such that each condition must occur a minimum number of times in each quartile of the training phase, thus resulting in an even distribution the conditions throughout training. While the assurance of the conditions being evenly spread throughout training is consistent across studies, other aspects of the sequence structure are a bit more idiosyncratic. Some researchers report setting a maximum number of consecutive repetitions, e.g. no more than 2 consecutive trials of the same condition [@delreyEffectsContextualInterference1982; @sheaContextualInterferenceEffects1979], or structure the random trials such that the same condition never occurs consecutively [@wulfEffectTypePractice1991]. Also common is to structure experiments such that random condition really consists of many small blocks, where participants do a few trials of one condition consecutively and then switch to another condition [@willeyLongtermMotorLearning2018; @chuaPracticeVariabilityPromotes2019; @wrisbergVariabilityPracticeHypothesis1987], resulting in many more switches than would arise if training was perfectly blocked. The question of whether such differences in the structure of random sequencing are consequential has been addressed experimentally a few times, in all cases consisting of a 1) a no-repeat random condition; 2) a blocked random condition (typically 3 or 4 repeats before a switch); and 3) a standard fully-blocked condition. Blocked-random training resulted in better performance than either repeat-random, or fully - blocked training in both a bean-bag throwing [@pigottMotorSchemaStructure1984], and basketball shot training study [@landinComparisonThreePractice1997], and in a replication plus extension of the seminal [@sheaContextualInterferenceEffects1979] study, blocked-random training was equally effective as no-repeat random training, with both random structures leading to better performance than the fully-blocked training condition. Consequences on different study schedules have also been repeatedly observed in the category learning literature [@carvalhoSequenceStudyChanges2017; @carvalhoPuttingCategoryLearning2014]. This line of research has revealed that the effects of blocking vs. interleaving can depend on the structure of the category being learned, and also that the different schedules can result in the participants requiring different representations. A fruitful line of inquiry in the motor skill learning literature may be to attempt to identify whether structural aspects of the motor task interact with different training sequences in a reliable manner.

Numerous researchers have attempted to provide coherent frameworks to account for the full range of influences of training variation and sequencing described above (along with many other effects not discussed). Such accounts are generally quite similar, invoking ideas of desirable levels of difficulty [@bjorkNewTheoryDisuse1992; @schmidtNewConceptualizationsPractice1992], or optimal challenge points [@guadagnoliChallengePointFramework2004]. They tend to start by describing the dissociation between acquisition performance (performance during training) and testing performance (delayed retention and/or transfer), most strikingly observed as varied/random training participants performing worse than their constant/blocked counterparts during the training stage of the study, but then outperforming the constant/blocked comparisons at a later retention or transfer stage. This observation is then used to justify the idea that the most enduring and generalizable learning occurs by training at an optimal level of training difficulty, with difficulty being some function of the experience of the learner, and the cognitive or visuomotor processing demands of the task. It then follows that the factors that tend to make training more difficult (i.e. increased variability or randomization), are more likely to be beneficial when the learner has some experience, or when the processing demands of the task are not too extreme (which may only occur after some experience with the task). Such frameworks may be helpful heuristics in some cases, but they also seem to be overly flexible such that any null result of some intervention might be accounted for by a suboptimal amount of training trials, or by suggesting the training task was too difficult. The development of computational models that can account for how changes in the parameters of the motor-skill task scale with difficulty, would be a great step forward.

>>>>

---
title: Introduction
format:
  html: default
  hugo-md:
    echo: false
    output: false
    html-math-method: mathjax
    output-file: d_intro_hugo.md
  gfm:
    echo: true
    output-file: d_into_gfm.md
  hikmah-manuscript-pdf:
    echo: false
    output-file: "intro_pdf.pdf"
    mainfont: "Linux Libertine O"
    mainfontoptions:
      - "Numbers=Proportional"
      - "Numbers=OldStyle"
    mathfont: "Libertinus Math"
---

@@

In the domain of category learning, the constant vs. varied comparison is much less suitable. Instead, researchers will 
typically employ designs where all training groups encounter numerous stimuli, but one group experiences a greater number of unique exemplars [@wahlheimMetacognitiveJudgmentsRepetition2012; @nosofskyModelguidedSearchOptimal2019; @doyleMetacognitiveMonitoringCategory2016; @hoschPriorExperienceVariability2023; @brunsteinPreparingNoveltyDiverse2011], or designs where the number of unique training exemplars is held constant, but one group trains with items that are more dispersed, or spread out across the category space [@posnerGenesisAbstractIdeas1968;  @homaCategoryBreadthAbstraction1976; @huHighvariabilityTrainingDoes2024; @bowmanTrainingSetCoherence2020; @maddoxStimulusRangeDiscontinuity2011].

Much of the earlier work in this sub-area trained subjects on artificial categories, such as dot patterns [@homaCategoryBreadthAbstraction1976; @posnerGenesisAbstractIdeas1968]. A seminal study by @posnerGenesisAbstractIdeas1968 trained participants to categorize artificial dot patterns, manipulating whether learners were trained with low variability examples clustered close to the category prototypes (i.e. low distortion training patterns), or higher-variability patterns spread further away from the prototype (i.e. high-distortion patterns). Participants that received training on more highly-distorted items showed superior generalization to novel high distortion patterns in the subsequent testing phase. It should be noted that unlike the sensorimotor studies discussed earlier, the @posnerGenesisAbstractIdeas1968 study did not present low-varied and high-varied participants with an equal number of training rathers, but instead had participants remain in the training stage of the experiment until they reached a criterion level of performance. This train-until-criterion procedure led to the high-variability condition participants tending to complete a larger number of training trials before switching to the testing stage. More recent work [@huHighvariabilityTrainingDoes2024], also used dot pattern categories, but matched the number of training trials across conditions. Under this procedure, higher-variability participants tended to reach lower levels of performance by the end of the training stage. The results in the testing phase were the opposite of @posnerGenesisAbstractIdeas1968,  with the low-variability training group showing superior generalization to novel high-distortion patterns (as well as generalization to novel patterns of low or medium distortion levels). However, whether this discrepancy is solely a result of the different training procedures is unclear, as the studies also differed in the nature of the prototype patterns used. @posnerGenesisAbstractIdeas1968  utilized simpler, recognizable prototypes (e.g., a triangle, the letter M, the letter F), while  @huHighvariabilityTrainingDoes2024 employed random prototype patterns.

Recent studies have also begun utilizing more complex or realistic sitmuli when assessing the influence of variability on category learning. @wahlheimMetacognitiveJudgmentsRepetition2012 conducted one such study. In a within-participants design, participants were trained on bird categories with either high repetitions of a few exemplars, or few repetitions of many exemplars. Across four different experiments, which were conducted to address an unrelated question on metacognitive judgements, the researchers consistently found that participants generalized better to novel species following training with more unique exemplars (i.e. higher variability), while high repetition training produced significantly better performance categorizing the specific species they had trained on. A variability advantage was also found in the relatively complex domain of rock categorization [@nosofskyModelguidedSearchOptimal2019]. For 10 different rock categories, participants were trained with either many repetitions of 3 unique examples of each category, or few repetitions of 9 unique examples, with an equal number of total training trials in each group (the design also included 2 other conditions less amenable to considering the impact of variation). The high-variability group, trained with 9 unique examples, showed significantly better generalization performance than the other conditions.

A distinct sub-literature within the category learning domain has examined how the variability or dispersion of the categories themselves influences generalization to ambiguous regions of the category space (e.g. the region between the two categories). The general approach is to train participants with examples from a high variability category and a low variability category. Participants are then tested with novel items located within ambiguous regions of the category space which allow the experimenters to assess whether the difference in category variability influenced how far participants generalize the category boundaries. @cohenCategoryVariabilityExemplar2001 trained subjects on two categories, one with much more variability than the other. In experiment 1, a low variability category composed of 1 instance was compared against a high-variability category of 2 instances in one condition, and 7 instances in another. In experiment 2 both categories were composed of 3 instances, but for the low-variability group the instances were clustered close to each other, whereas the high-variability groups instances were spread much further apart. Participants were tested on an ambiguous novel instance that was located in between the two trained categories. Both experiments provided evidence that participants were much more likely to categorize the novel middle stimulus into a category with greater variation.

Further observations consonant with the results of @cohenCategoryVariabilityExemplar2001  have since been observed in numerous investigations [@hahnEffectsCategoryDiversity2005; @perlmanFurtherAttemptsClarify2012;
@sakamotoPuttingPsychologyBack2008; @hsuEffectsGenerativeDiscriminative2010; but see @stewartEffectCategoryVariability2002]. The results of @sakamotoPuttingPsychologyBack2008 are noteworthy. They first reproduced the basic finding of participants being more likely to categorize an unknown middle stimulus into a training category with higher variability. In a second experiment, they held the variability between the two training categories constant and instead manipulated the training sequence, such that the examples of one category appeared in an ordered fashion, with very small changes from one example to the other (the stimuli were lines that varied only in length), whereas examples in the alternate category were shown in a random order and thus included larger jumps in the stimulus space from trial to trial. They found that the middle stimulus was more likely to be categorized into the category that had been learned with a random sequence, which was attributed to an increased perception of variability which resulted from the larger trial to trial discrepancies.

The work of @hahnEffectsCategoryDiversity2005, is also of particular interest to the present work. Their experimental design was similar to previous studies, but they included a larger set of testing items which were used to assess generalization both between the two training categories as well as novel items located in the outer edges of the training categories. During generalization testing, participants were given the option to respond with "neither", in addition to responses to the two training categories. The "neither" response was included to test how far away in the stimulus space participants would continue to categorize novel items as belonging to a trained category. Consistent with prior findings, high-variability training resulted in an increased probability of categorizing items in between the training categories as belong to the high variability category. Additionally, participants trained with higher variability also extended the category boundary further out into the periphery than participants trained with a lower variability category were willing to do. The author compared a variety of similarity-based models based around the Generalized Context Model [@nosofskyAttentionSimilarityIdentificationcategorization1986] to account for their results, manipulating whether a response-bias or similarity-scaling parameter was fit separately between variability conditions. No improvement in model fit was found by allowing the response-bias parameter to differ between groups, however the model performance did improvement significantly when the similarity scaling parameter was fit separately. The best fitting similarity-scaling parameters were such that the high-variability group was less sensitive to the distances between stimuli, resulting in greater similarity values between their training items and testing items. This model accounted for both the extended generalization gradients of the varied particpants, and also for their poorer performance in a recognition condition.

Variability has also been examined in the learning of higher-order linguistic categories [@perryLearnLocallyThink2010]. In nine training sessions spread out over nine weeks infants were trained on object labels in a naturalistic play setting. All infants were introduced to three novel objects of the same category, with participants in the "tight" condition being exposed to three similar exemplars of the category, and participants in the varied condition being exposed to three dissimilar objects of the same category. Importantly, the similarity of the objects was carefully controlled for by having a separate group of adult subjects provide pairwise similarity judgements of the category objects prior to the study onset. Multidimensional scaling was then performed to obtain the coordinates of the objects psychological space, and out of the 10 objects for each category, the 3 most similar objects were selected for the tight group and the three least similar objects for the varied group, with the leftover four objects being retained for testing. By the end of the nine weeks, all of the infants had learned the labels of the training objects. In the testing phase, the varied group demonstrated superior ability to correctly generalize the object labels to untrained exemplars of the same category. More interesting was the superior performance of the varied group on a higher order generalization task -- such that they were able to appropriately generalize the bias they had learned during training for attending to the shape of objects to novel solid objects, but not to non-solids. The tight training group, on the other hand, tended to overgeneralize the shape bias, leading the researchers to suggest that the varied training induced a more context-sensitive understanding of when to apply their knowledge.

Of course, the relationship between training variability and transfer is unlikely to be a simple function wherein increased variation is always beneficial. Numerous studies have found null, or in some cases negative effects of training variation [@deloshExtrapolationSineQua1997; @sinkeviciuteRoleInputVariability2019; @wrisbergVariabilityPracticeHypothesis1987; @vanrossumSchmidtSchemaTheory1990], and many more have suggested that the benefits of variability may depend on additional factors such as prior task experience, the order of training trials, or the type of transfer being measured [@bernikerEffectsTrainingBreadth2014; @braithwaiteEffectsVariationPrior2015; @hahnEffectsCategoryDiversity2005; @lavanEffectsHighVariability2019; @northEffectConsistentVaried2019; @sadakataIndividualAptitudeMandarin2014; @zamanPerceptualVariabilityImplications2021].

In an example of a more complex influence of training variation, [@braithwaiteEffectsVariationPrior2015] trained participants on example problems involving the concept of sampling with replacement (SWR). Training consisted of examples that were either highly similar in their semantic context (e.g. all involving people selecting objects) or in which the surface features were varied between examples (e.g. people choosing objects AND objects selected in a sequence). The experimenters also surveyed how much prior knowledge each participant had with SWR. They found that whether variation was beneficial depended on the prior knowledge of the participants -- such that participants with some prior knowledge benefited from varied training, whereas participants with minimal prior knowledge performed better after training with similar examples. The authors hypothesized that in order to benefit from varied examples, participants must be able to detect the structure common to the diverse examples, and that participants with prior knowledge are more likely to be sensitive to such structure, and thus to benefit from varied training. To test this hypothesis more directly, the authors conducted a 2nd experiment, wherein they controlled prior knowledge by exposing some subjects to a short graphical or verbal pre-training lesson, designed to increase sensitivity to the training examples. Consistent with their hypothesis, participants exposed to the structural sensitivity pre-training benefited more from varied training than the controls participants who benefited more from training with similar examples. Interactions between prior experience and the influence of varied training have also been observed in sensorimotor learning [@guadagnoliRelationshipContextualInterference1999;  @delreyEffectsContextualInterference1982].  @delreyEffectsContextualInterference1982 recruited participants who self-reported either extensive, or very little experience with athletic activities, and then trained participants on a coincident timing task under with either a single constant training velocity, with one of several varied training procedures. Unsurprisingly, athlete participants had superior performance during training, regardless of  condition, and training performance was superior for all subjects in the constant group. Of greater interest is the pattern of testing results from novel transfer conditions. Among the athlete-participants, transfer performance was best for those who received variable training. Non-athletes showed the opposite pattern, with superior performance for those who had constant training.

### The current work

The overarching purpose of this dissertation is to investigate the effects of training variability on learning and generalization within visuomotor skill learning and function learning. Our investigations is structured into two main projects,  each employing distinct experimental paradigms and computational modeling frameworks to elucidate how and when variability in training enhances or impedes subsequent generalization.

In Project 1, we investigated the influence of varied practice in a simple visuomotor projectile launching task. Experiments 1 and 2 compared the performance of constant and varied training groups to assess potential benefits of variability on transfer to novel testing conditions. To account for the observed empirical effects, we introduced the Instance-based Generalization with Adaptive Similarity (IGAS) model. IGAS provides a novel computational approach for quantifying the similarity between training experiences and transfer conditions, while also allowing for variability to influence the generalization gradient itself.

Project 2 shifted focus to the domain of function learning by employing a visuomotor extrapolation task. Across three experiments, we examined how constant and varied training regimes affected learning, discrimination between stimuli, and the ability to extrapolate to novel regions of the function's input space. To model human performance in this task, we fit the influential Associative Learning Model (ALM) and the Extrapolation-Association Model (EXAM) to individual participant data using advanced Bayesian parameter estimation techniques.
